k-plus Anticlustering: An Improved k-means Criterion for Maximizing Between-Group Similarity

### 2023-07-04 paper.Rmd
do not use mathbold for scalars

Note that there is a direct connection between the $\mathit{SSE}$ and the location of the cluster centroids [@spath1986]. If the $\mathit{SSE}$ is minimal---which is the goal of k-means *clustering*---, the centroid values $\overline{\mathbf{x}}^{(k)} (k = 1, \ldots, K)$ are as far away as possible from the overall data centroid $\overline{\mathbf{x}} = (\overline{\mathbf{x}}_1, \overline{\mathbf{x}}_2, \ldots, \overline{\mathbf{x}}_P$), where $\overline{\mathbf{x}}_j$ ($j = 1, \ldots, P$) is the overall mean value on the $j$'th attribute:

$$
\overline{\mathbf{x}}_j = \frac{1}{N} \sum\limits_{i=1}^{N} x_{ij}
$$

@@

Thus, to obtain an adaptation of the $\mathit{SSE}$ whose maximization can minimize differences in variances between groups, a new variable is computed for each original variable: The new variable represents the squared distance of each of the original values to the overall mean of the variable. We obtain a new data matrix $\mathbf{Y}_{N \times P} = \{y_{ij}\}_{N \times P}$ with $y_{ij} = (x_{ij} - \overline{\mathbf{x}}_j)^2$. On the basis of $\mathbf{Y}_{N \times P}$, we define the criterion $\mathit{SSE_{Var}}$:

@@

$$
\frac{1}{N} \sum\limits_{i=1}^{N} (x_i - \overline{\mathbf{x}})^t.
$$

@@

We define the $j$'th column ($j = 1, \ldots, P$) of the original data matrix $\mathbf{X}_{N \times P}$ as $x_{\cdot j} =  (x_{1j}, \ldots, x_{Nj})$ where the column mean---the overall mean of the $j$'th attribute---is $\overline{\mathbf{x}}_j$. Let $T$ be the number of distribution moments that should be similar across $K$ groups. For each moment $t$ ($t = 1, \ldots, T$) and for each variable $x_{\cdot j}$ ($j = 1, \ldots, P$), a new *k-plus variable* $\hat{x}_{t \cdot j} = (\hat{x}_{t1j}, \ldots, \hat{x}_{tNj})$ is defined where

$$
\hat{x}_{tij} = 
\begin{cases}
    x_{ij} & \text{if } t = 1\\
    (x_{ij} - \overline{\mathbf{x}}_j)^t, & \text{if } t > 1.
\end{cases}
$$

@@

$$
  \mathbf{X'}_{N \times \mathit{TP}} =
  \left[ {\begin{array}{ccc|ccc|c|ccc}
    x_{11} & \cdots & x_{1P} & 
        (x_{11} - \overline{\mathbf{x}}_1)^2 & \cdots & (x_{1P} - \overline{\mathbf{x}}_P)^2 & 
        \cdots & 
        (x_{11} - \overline{\mathbf{x}}_1)^T & \cdots & (x_{1P} - \overline{\mathbf{x}}_P)^T \\
    x_{21} & \cdots & x_{2P} & 
        (x_{21} - \overline{\mathbf{x}}_1)^2 & \cdots & (x_{2P} - \overline{\mathbf{x}}_P)^2 & 
        \cdots & 
        (x_{21} - \overline{\mathbf{x}}_1)^T & \cdots & (x_{2P} - \overline{\mathbf{x}}_P)^T \\
    \vdots & \ddots & \vdots & \vdots & \ddots & \vdots & \cdots & \vdots & \ddots & \vdots\\
    x_{N1} & \cdots & x_{NP} & 
        (x_{N1} - \overline{\mathbf{x}}_1)^2 & \cdots & (x_{NP} - \overline{\mathbf{x}}_P)^2 & 
        \cdots & 
        (x_{N1} - \overline{\mathbf{x}}_1)^T & \cdots & (x_{NP} - \overline{\mathbf{x}}_P)^T \\
  \end{array} } \right]
$$

>>>>

Note that there is a direct connection between the $\mathit{SSE}$ and the location of the cluster centroids [@spath1986]. If the $\mathit{SSE}$ is minimal---which is the goal of k-means *clustering*---, the centroid values $\overline{\mathbf{x}}^{(k)} (k = 1, \ldots, K)$ are as far away as possible from the overall data centroid $\overline{\mathbf{x}} = (\overline{x}_1, \overline{x}_2, \ldots, \overline{x}_P)$, where $\overline{x}_j$ ($j = 1, \ldots, P$) is the overall mean value on the $j$'th attribute:

$$
\overline{x}_j = \frac{1}{N} \sum\limits_{i=1}^{N} x_{ij}
$$

@@

Thus, to obtain an adaptation of the $\mathit{SSE}$ whose maximization can minimize differences in variances between groups, a new variable is computed for each original variable: The new variable represents the squared distance of each of the original values to the overall mean of the variable. We obtain a new data matrix $\mathbf{Y}_{N \times P} = \{y_{ij}\}_{N \times P}$ with $y_{ij} = (x_{ij} - \overline{x}_j)^2$. On the basis of $\mathbf{Y}_{N \times P}$, we define the criterion $\mathit{SSE_{Var}}$:

@@

$$
\frac{1}{N} \sum\limits_{i=1}^{N} (x_i - \overline{x})^t.
$$

@@

We define the $j$'th column ($j = 1, \ldots, P$) of the original data matrix $\mathbf{X}_{N \times P}$ as $x_{\cdot j} =  (x_{1j}, \ldots, x_{Nj})$ where the column mean---the overall mean of the $j$'th attribute---is $\overline{x}_j$. Let $T$ be the number of distribution moments that should be similar across $K$ groups. For each moment $t$ ($t = 1, \ldots, T$) and for each variable $x_{\cdot j}$ ($j = 1, \ldots, P$), a new *k-plus variable* $\hat{x}_{t \cdot j} = (\hat{x}_{t1j}, \ldots, \hat{x}_{tNj})$ is defined where

$$
\hat{x}_{tij} = 
\begin{cases}
    x_{ij} & \text{if } t = 1\\
    (x_{ij} - \overline{x}_j)^t, & \text{if } t > 1.
\end{cases}
$$

@@

$$
  \mathbf{X'}_{N \times \mathit{TP}} =
  \left[ {\begin{array}{ccc|ccc|c|ccc}
    x_{11} & \cdots & x_{1P} & 
        (x_{11} - \overline{x}_1)^2 & \cdots & (x_{1P} - \overline{x}_P)^2 & 
        \cdots & 
        (x_{11} - \overline{x}_1)^T & \cdots & (x_{1P} - \overline{x}_P)^T \\
    x_{21} & \cdots & x_{2P} & 
        (x_{21} - \overline{x}_1)^2 & \cdots & (x_{2P} - \overline{x}_P)^2 & 
        \cdots & 
        (x_{21} - \overline{x}_1)^T & \cdots & (x_{2P} - \overline{x}_P)^T \\
    \vdots & \ddots & \vdots & \vdots & \ddots & \vdots & \cdots & \vdots & \ddots & \vdots\\
    x_{N1} & \cdots & x_{NP} & 
        (x_{N1} - \overline{x}_1)^2 & \cdots & (x_{NP} - \overline{x}_P)^2 & 
        \cdots & 
        (x_{N1} - \overline{x}_1)^T & \cdots & (x_{NP} - \overline{x}_P)^T \\
  \end{array} } \right]
$$
